Title of Paper

### 2017-02-17 paper.tex
move data access / vec patterns discussion up to intro to save space in discussion of past work; update plots

\begin{figure}[htb]
	\centering
	\includegraphics[width=0.33\linewidth]{SIMD.pdf}
	\caption{Schematic of SIMD processing.  A single processing element (e.g. a CPU core) contains a vector unit with several lanes (L0, L1, etc.).  The vector unit executes a single instruction concurrently on multiple data.}
	\label{F:SIMD}
\end{figure}

@@

Several recent works, e.g. \cite{CurtisGPU:2017,Sewerin20151375} have investigated SIMT-based chemical kinetic integration on GPUs, however SIMD-based chemical kinetics have been less studied in comparison.
\textcite{stone2016} implemented a linearly-implicit fourth-order stiff Rosenbrock solver in the OpenCL for various platforms including the CPU, GPU and the MIC; the Rosenbrock solver performance was compared to that of a non-stiff, fourth-order explicit Runge–Kutta–Fehlberg integrator.
Two parallelization models were investigated in OpenCL, the so called \textit{shallow}-SIMD vectorization, where each SIMD-lane solved a single chemical kinetic initial value problem (IVP), and a SIMT-vectorization where each thread solved a single IVP.
Shallow-SIMD vectorization improved the integrator performance over the OpenMP baseline by \SIrange{2.5}{2.8}{$\times$} on the CPU and \SIrange{4.7}{4.9}{$\times$} on the MIC, while the GPU performance while the GPU was \SIrange{1.4}{1.6}{$\times$} slower than the OpenMP baseline due to thread-divergence concerns.
\textcite{kroshko2013efficient} implemented a shallow-SIMD -vectorized third order Rosenbrock solver (RODAS) integrator for atmospheric chemistry on a Cell Broadband Engine---a specially designed vector processor---finding a speedup of \SI{1.89}{$\times$} over a serial version of the same code, a \SI{94}{\percent} parallel efficiency.

This work will study the performance of chemical source term evaluation of automatically generated shallow-SIMD-vectorized codes for a wide range chemical kinetic models on the CPU.
The performance will be compared to a baseline SIMT-vectorized codes to determine the effective SIMD speedup.
Finally future extensions to this work will be detailed.

@@

\subsection{Data Ordering and Vectorization Patterns}
\label{S:data}
When storing arrays for a chemical kinetic model, e.g. the concentration of species $k$ for the $j$-th of thermo-chemical state $[C_k]_j$ there are two data-ordering choices.
In the ``F'' (Fortran column-major) format, the concentrations of a single species $k$ over all $N$ thermo-chemical states are adjacent: $[C_k]_j, [C_k]_{j + 1}\ldots[C_k]_{N}$;
conversely in the ``C'' (C row-major) format, the species concentrations for the $j$-th thermo-chemical state: $[C_k]_j, [C_{k+1}]_j\ldots$ are stored sequentially in memory;
This ordering---along with the device (CPU, GPU, etc.) in question---have a large effect on the performance of SIMD\slash SIMT-vectorized algorithms.

In this work a \textit{shallow}-SIMD vectorization, in which each SIMD-lane evaluates chemical source terms for a different thermo-chemical state, has been implemented.
If the data is stored in ``F''-order, the SIMD-lanes will access sequential locations in memory however the next species' concentration---for example---will be $N$ memory locations away, potentially resulting in more cache-misses on the CPU.
In future efforts, a so called \textit{deep}-SIMD vectorization will be pursued, in which a vector-unit cooperates to evaluate chemical source terms for a single thermo-chemical state.
In this context, a deep-vectorization may result in SIMD-\textit{waste}---similar to thread-divergence in GPU-processing---caused by different SIMD-lanes executing different instructions (e.g. from differing for-loop bounds).
However, SIMD-lanes in a ``C''-ordered deep-vectorization access sequential memory locations and further the data locality of e.g., species concentrations, will likely result in improved CPU cache-hit rates.
Past studies additionally demonstrate promising results for deep-SIMT vectorization on GPUs~\cite{Sewerin20151375}.
In Sec.~\ref{S:results} the effect of this ordering on the shallow-SIMD vectorization performance will be studied.

>>>>

\begin{figure}[htb]
  \centering
  \begin{subfigure}[t]{0.54\linewidth}
      \includegraphics[width=\textwidth]{SIMD.pdf}
      \caption{Schematic of SIMD processing.  A single processing element (e.g. a CPU core) contains a vector unit with several lanes (L0, L1, etc.).  The vector unit executes a single instruction concurrently on multiple data.}
      \label{F:SIMD}
  \end{subfigure}
  \hfill
  \begin{subfigure}[t]{0.43\linewidth}
      \includegraphics[width=\textwidth]{mem_access.pdf}
      \caption{Representation of an 2-D array with dimension $N_1 \times N_2$ in CPU memory, with linear index marked.}
      \label{F:mem}
  \end{subfigure}
\end{figure}

@@

\subsection{Data ordering and vectorization patterns}
\label{S:data}
When storing arrays for a chemical kinetic model, e.g. the concentration of species $k$ for the $j$-th of thermo-chemical state $[C_k]_j$ there are two data-ordering choices.
In the ``F'' (Fortran column-major) format, the concentrations of a single species $k$ over all $N$ thermo-chemical states are adjacent.
In Fig.~\ref{F:mem} this would correspond to storing $[C_0]_0$ in index \num{0}, $[C_0]_1$ in index \num{1} and so on, with $N_1 = N$, and $N_2 = N_s$, the number of species in the model.
In the ``C'' (C row-major) format, the storage is flipped with $[C_0]_0$ in index \num{0}, $[C_1]_0$ in index \num{1} etc., with $N_1 = N_s$ and $N_2 = N$.
This ordering---along with the device (CPU, GPU, etc.) in question---have a large effect on the performance of SIMD\slash SIMT-vectorized algorithms.

In a \textit{shallow}-SIMD vectorization, each SIMD-lane evaluates chemical source terms for a different thermo-chemical state, has been implemented.
If the data is stored in ``F''-order, the SIMD-lanes accessing the concentration of species $k$ for different states will load sequential memory locations in Fig.~\ref{F:mem} in memory however the $k+1$'th species concentration will be $N$ memory locations away, increasing the likelihood of cache-misses on the CPU.
In a \textit{deep}-SIMD vectorization, a vector-unit cooperates to evaluate chemical source terms for a single thermo-chemical state, thus SIMD-lanes loading species concentrations $k\dots k+V$ (for a vector width of $V$) will access sequential memory locations if the data is stored in ``C''-order.
Further, in the ``C''-ordering the furthest difference between any two concentrations in a single thermo-chemical state is at most $N_s \ll N$ (in most cases), greatly improving the data-locality and chance of a cache-hit on the CPU.
However, a deep-vectorization may result in SIMD-\textit{waste}---similar to thread-divergence in GPU-processing---caused by different SIMD-lanes executing different instructions (e.g. from differing for-loop bounds).
In Sec.~\ref{S:results} the effect of this ordering on the shallow-SIMD vectorization performance will be studied.

\subsection{Previous works and goals of this tudy}
Several recent works, e.g. \cite{CurtisGPU:2017} have investigated SIMT-based chemical kinetic integration on GPUs, however SIMD-based chemical kinetics have been less studied in comparison.
\textcite{stone2016} implemented a linearly-implicit fourth-order stiff Rosenbrock solver in the OpenCL for various platforms including the CPU, GPU and the MIC.
The shallow-SIMD and SIMT-vectorization models were investigated in OpenCL and compared to an OpenMP baseline code that was deep-vectorized by simple compiler hints (a.k.a pragmas).
The shallow vectorization improved the integrator performance over the OpenMP baseline by \SIrange{2.5}{2.8}{$\times$} on the CPU and \SIrange{4.7}{4.9}{$\times$} on the MIC, while the GPU performance while the GPU was \SIrange{1.4}{1.6}{$\times$} slower than the OpenMP baseline due to thread-divergence concerns.
\textcite{kroshko2013efficient} implemented a shallow-SIMD vectorized third order stiff Rosenbrock integrator for atmospheric chemistry on a Cell Broadband Engine---a specially designed vector processor---finding a speedup of \SI{1.89}{$\times$} over a serial version of the same code, a \SI{94}{\percent} parallel efficiency.

This work will study the performance of chemical source term evaluation of automatically generated shallow-SIMD vectorized codes for a wide range chemical kinetic models on the CPU.
The performance will be compared to a baseline SIMT vectorized code to determine the effective SIMD speedup.
Finally future extensions to this work will be detailed.

@@
